---
title: "R Solo Project: School Districts"
output: html_notebook
---

# Imports

```{r}
library(tidyverse)
```


# Exercises

## Problem 1

Read in districts.csv into a tibble named districts. How many rows and columns does it contain?

```{r}
districts <- read_csv('../data/districts.csv')
str(districts)
```

### Answer 1
Rows: 146 Columns: 27

## Problem 2

Notice that the first row corresponds to the whole State of Tennessee. Remove this row and save the result back to districts.

```{r}
districts <- districts %>%
  filter(system_name != "State of Tennessee")
```

## Problem 3

How many districts have a proficiency rate of at least 80% for both alg_1 and eng_1?

```{r}
districts %>%
  filter(alg_1 >= 80 & eng_1 >= 80)
```
### Answer 3

13 school districts

4. How many districts have a proviciency rate less than 50% for either alg_1 or eng_1?

```{r}
districts %>%
  filter(alg_1 < 50 | eng_1 < 50)
```
### Answer 4

8 districts

## Problem 5

Which district has the lowest graduation rate?

```{r}
districts %>%
  slice_min(grad, n = 1)
```
### Answer 5
The district with the lowest graduation rate is the Tennessee School for the Blind with 11.1%.

## Problem 6

Which district has the highest dropout rate?

```{r}
districts %>%
  slice_max(dropout, n = 1)
```
### Answer 6

The district with the highest dropout rate is the Achievement School District with 27.6%.

## Problem 7

Within the Mid Cumberland region, which district has the highest ACT composite?

```{r}
districts %>%
  filter(region == "Mid Cumberland") %>%
  slice_max(act_composite, n = 1)
```
### Answer 7

Williamson County has the highest ACT composite in the Mid Cumberland region

## Problem 8

Create a histogram showing the distribution of graduation rates. What can you say about this distribution?

```{r}
districts %>%
  ggplot(aes(x = grad)) +
    geom_histogram() +
    ggtitle("Distribution of Graduation Rates") +
    xlab("Graduation Rate")
```
```{r}
districts %>%
  ggplot(aes(x = grad)) +
    geom_histogram() +
    ggtitle("Distribution of Graduation Rates") +
    xlab("Graduation Rate") +
    facet_wrap(~ region)
```


### Answer 8
The distribution is heavily left skewed, with a few low outliers.

## Problem 9

Create a scatter plot to compare alg_1 proficiency rates to alg_2 rates. What do you notice? Facet this plot by region. Does anything stand out when you facet the plots?

```{r}
districts %>%
  ggplot(aes(x = alg_1, y = alg_2)) +
    geom_point() +
    facet_wrap(~ region)
```
### Answer 9

alg_1 and alg_2 tend to be positively correlated.The exceptions appear to be Southeast and Upper Cumberland.

## Problem 10

Create a bar chart showing the total enrollment by region. Which region has the highest total enrollment? Which has the smallest?

```{r}
districts %>%
  group_by(region) %>%
  summarize(total_enrollment = sum(enrollment, na.rm = TRUE)) %>%
  mutate(regionfct = factor(region)) %>%
  mutate(regionfct = fct_reorder(regionfct, total_enrollment, .desc = FALSE)) %>%
  ggplot(aes(y = regionfct , x = total_enrollment, fill = region)) +
    geom_col() +
    labs(title = "Total Enrollment by Region",
         x = "Total Enrollment",
         y = "Region")
```
### Answer 10
Mid Cumberland has the highest total enrollment; Northwest has the lowest total enrollment.

## Problem 11

When creating this bar chart you may have noticed that some districts have missing enrollment values. For how many districts is this the case?

```{r}
districts %>%
  group_by(region) %>%
  summarize(na_count = sum(is.na(enrollment))) %>%
  arrange(desc(na_count), region)
```

### Answer 11
The regions with an NA are East TN, Mid Cumberland, Southwest/Memphis, and Upper Cumberland.

## Problem 12

What is the mean graduation rate across all districts? What might be wrong with using just the regular mean to assess average graduation rates?

```{r}
districts %>%
  summarize(mean_grad = mean(grad, na.rm = TRUE))
```
### Answer 12
The mean graduation rate across all districts is about 90.1%. Using the regular mean to assess graduation rates assumes that all districts should have equal weight in calculating the mean; however, some districts might represent more or less students than others.

## Problem 13

Redo the previous question but use a weighted average (weighted.mean) graduation across all districts, weighing by enrollment. How much does this change your answer? Can you explain using the data the reason for the big change from using the mean?

```{r}
districts %>%
  summarize(weighted_mean_grad = weighted.mean(grad, coalesce(enrollment,0), na.rm = TRUE))
```
```{r}
districts %>%
  group_by(region) %>%
  summarize(region_weighted_grad = weighted.mean(grad, coalesce(enrollment, 0), na.rm = TRUE)) %>%
  arrange(desc(region_weighted_grad))

```

### Answer 13
The weighted mean graduation rate across all districts is about 87.3%, which is nearly 3% lower than the regular mean. This is not surprising given that the three regions with the highest enrollment are also three of the four with the lowest weighted mean graduation rates.

## Problem 14

Create a boxplot showing enrollment rates per region. Does anything stand out?

```{r}
districts %>%
  ggplot(aes(y = region, x = enrollment, fill = region)) +
    geom_boxplot()
  
```

### Answer 14
Enrollments of school districts are right-skewed. With the exception of Northwest and First TN, each region has at least one outlier district with unusually high enrollment (and Southwest/Memphis with one that's just ridiculous.) Also, Mid Cumberland has by far the most variation, including in its IQR.


## Problem 15

Create a boxplot showing graduation rates per region. Does anything stand out?

```{r}
districts %>%
  group_by(region) %>%
  ggplot(aes(y = region, x = grad, fill = region)) +
    geom_boxplot()
```

### Answer 15

Graduation rates of school districts are largely left-skewed or normal with low outliers. Mid Cumberland has a ridiculously low outlier (TN School for the Blind).Southwest/Memphis has the highest variation, including in its IQR.

## Problem 16

Find the weighted average of graduation rates by region using enrollment as weights. Compare the results you get for the weighted average to what you see from the boxplots. Can you explain any discrepancy? (Did this in Problem 13, as well.)

```{r}
districts %>%
  group_by(region) %>%
  summarize(region_weighted_grad = weighted.mean(grad, coalesce(enrollment, 0), na.rm = TRUE)) %>%
  arrange(desc(region_weighted_grad))
```
### Answer 16
*We'll come back to this.*

## Problem 17

For many districts, values for alg_2 are lower than for alg_1. Create a histogram showing the distribution of differences (alg_1 - alg_2). Which school had the largest drop from alg_1 to alg_2? For what percentage of schools is it true that alg_2 is larger than alg_1? Is there a similar dropoff for eng_2 and eng_3?

```{r}
districts_algebra <- districts %>%
  summarize(alg_diff = alg_1 - alg_2)
  
ggplot(districts_algebra, aes(x = alg_diff, fill = alg_diff < 0)) +
  geom_histogram() +
  theme(legend.title=element_blank()) +
  scale_fill_discrete(name = "", 
                      labels = c("Scores Drop", "Scores Rise"), 
                      guide = guide_legend(reverse=TRUE)) +
  labs(title = "Histogram of Differences between Algebra 1 and Algebra 2",
       x = "Difference between Algebra 1 and Algebra 2 \n (Positive values represent a drop from Algebra 1 to Algebra 2)")
    
```


18. You may have noticed that a lot of rows are missing values. Which district has the largest number of missing values? What do you notice about schools that have a lot of missing values?

```{r}

```


19. Find the correlation between graduation rate and all other variables. Create a horizontal bar chart showing these correlations. Make sure that your plot is ordered by correlation values. What do you notice from these correlations?

```{r}

```


20. Create a scatterplot for grad vs. suspended. Does what you see make sense given your answer from the previous part?

```{r}

```


21. Create a linear regression model using lm with target variable grad and predictor variable suspended. What R^2 value does this model have? What is the interpretation of this number?

```{r}

```


22. Add the regression line to your scatterplot using geom_smooth with method='lm'. How do you feel about the regression line after seeing it plotted on the scatterplot?

```{r}

```


Continued Exploration and Practice

23. Read in the school-level testing data for 2014, available here. You might find the readxl library useful for this task. If you use this library, be sure to look at the na argument for the read_excel function.

```{r}

```


24. How many schools have at least 20 percent of students below bsc for Algebra I? Which districts do these schools belong to?

```{r}

```


25. How many schools have at least 20 percent of students below bsc for both Algebra I and English I?

```{r}

```


26. Which grade has the highest pct_adv for Algebra I? Plot the average pct_adv per grade level as a bar chart. Make sure that the bars are ordered by grade level.

```{r}

```


27. Find the correlation between pct_adv for Algebra I and pct_adv for Algebra II by school. Create a scatterplot showing Algebra II scores vs. Algebra I scores by school.

```{r}

```


28. Find all schools in Rutherford County that have "High School" in their name. For these schools, create a chart (your choice) showing the differences in pct_below_bsc, pct_bsc, pct_prof, and pct_adv for Algebra I when looking across all subgroups and grades.

```{r}

```


29. I claim that smaller schools do a better job preparing students for Algebra I standardized tests. Find the average number of valid tests (a proxy for the school size) for schools where the pct_prof_adv for Algebra I is greater than 95. Compare this to the average number of valid tests for all schools. In light of this result, how does my claim look?

```{r}

```


30. I also claim that smaller schools do a worse job preparing students for Algebra I standardized tests. Find the average number of valid tests (a proxy for the school size) for schools where the pct_prof_adv for Algebra I is less than 25. Compare this to the average number of valid tests for all schools. In light of this result, how does my claim look now?

```{r}

```


31. Create a scatterplot showing pct_prov_adv vs. valid_tests. Can you use this to explain the result for numbers 26 and 27?

```{r}

```


If you finish all of the above questions, continue to explore the two datasets and see what else interesting you can find.

Also, check out the plotly library for R. The ggplotly function makes it very easy to convert ggplot plots into interactive plotly plots.